%%% Abstract in English

Static analysis aims to achieve an understanding of program behavior,
by means of automatic reasoning that requires only the program's
source code and not any actual execution. To reach a truly broad level
of program understanding, static analysis techniques need to create an
abstraction of memory that covers all possible executions. Such
abstract models may quickly degenerate after losing essential
structural information about the memory objects they describe, due to
the use of specific programming idioms and language features, or
because of practical analysis limitations. In many cases, some of the
lost memory structure may be retrieved, though it requires complex
inference that takes advantage of indirect uses of types. Such
recovered structural information may, then, greatly benefit static
analysis.

This dissertation shows how we can recover structural information,
first
\begin{inparaenum}[(i)]
\item in the context of C/C++, and next, in the context of
  higher-level languages without direct memory access, like Java,
  where we identify two primary causes of losing memory structure:
\item the use of reflection, and
\item analysis of partial programs.
\end{inparaenum}
We show that, in all cases, the recovered structural information
greatly benefits static analysis on the program.

For C/C++, we introduce a structure-sensitive pointer analysis that
refines its abstraction based on type information that it discovers
on-the-fly. This analysis is implemented in \cclyzer{}, a static
analysis tool for LLVM bitcode.
%
Next, we present techniques that extend a standard Java pointer
analysis by building on top of state-of-the-art handling of
reflection. The principle is similar to that of our
structure-sensitive analysis for C/C++: track the use of reflective
objects, \emph{during} pointer analysis, to gain important insights on
their structure, which can be used to ``patch'' the handling of
reflective operations on the running analysis, in a mutually recursive
fashion.
%
Finally, to address the challenge of analyzing partial Java programs
in full generality, we define the problem of ``program
complementation'': given a partial program we seek to provide
definitions for its missing parts so that the ``complement'' satisfies
all static and dynamic typing requirements induced by the code under
analysis. Essentially, complementation aims to recover the structure
of phantom types.  Apart from discovering missing class members (i.e.,
fields and methods), satisfying the subtyping constraints leads to the
formulation of a novel typing problem in the OO context, regarding
type hierarchy complementation. We offer algorithms to solve this
problem in various inheritance settings, and implement them in
JPhantom, a practical tool for Java bytecode complementation.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% End:
