I would like to express my deepest gratitude to my advisor, Yannis
Smaragdakis, for being such a great mentor to me, throughout my PhD
studies. I immensely appreciate his endless encouragement, patience,
and motivation; his advice and guidance have been priceless. I hope
that I can have such a huge impact on a person, as he had on me.

I also thank Alex Delis, Panos Rondogiannis, Mema Roussopoulos,
Manolis Koubarakis, Nikos Papaspyrou, and Kostis Sagonas for their
valuable comments and advice while serving as members of my
dissertation committee.

My sincere thanks to Shan Shan Huang, Martin Bravenboer, and Molham
Aref, who gave me the opportunity to join LogicBlox as an intern,
during my PhD. It was a wonderful experience and I'm indebted to all
those who made it happen.

A special thanks to my labmates and colleagues: Aggelos Biboudis,
Kostas Ferles, George Kollias, George Kastrinis, George Fourtounis,
Neville Grech, Anastasis Antoniadis, Efthymios Hadjimichael, Petros
Pathoulas, Dimitris Galipos, Stamatis Kolovos, Konstantinos
Triantafyllou, and Kostas Saidis.
%
I'm grateful for our interactions, discussions, and arguments over
every possible aspect of programming languages and software
engineering that intrigued us, and for all the fun we had.

I thank Sofia for standing by me in all times and always being
positive and understanding.

Finally, I would like to thank my parents, Dimitris and Aggeliki, for
being so supportive and reassuring over all these years.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% End:
